---
title: "mrr_project_bio"
author: ""
date: "2025-11-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Classification of Cancer outcome using Genetic and Clinical data

## Introduction

This project investigates the effect of genetic and clinical variables on the survival outcome of breast cancer patients. The data contains records of more than 1000 breast cancer patients from several research institutions. Clinical data contains patient-related and tumor-related information. Additionally mRNA gene expression data is available for each patient. The gene expression data has been processed to include only the top 5000 most variable genes on the transformed scale (log2(counts + 1)).

We consider that the main outcome variable of interest is **vital_status**, which is defined in clinical data.

## Load data

```{r}
load("mrr_bio.Rdata")
```

### Genes data

```{r}
dim(GeneX)
str(GeneX)
colnames(GeneX)[1:10]
rownames(GeneX)[1:10]
```

`GeneX` is a table with 1231 rows and 5000 columns:

- The rows correspond to the patients with each row name being the patient identifier such as `EW-A2FS-01A-11R-A17B`.
- The columns correspond to the genes with each column name being a gene such as `CLEC3A`, which are the 5000 most variable genes from one person to another.
- Each cell corresponds to the expression value (a float) of a certain gene (column) for a certain patient (row) -> continuous variables

### Clinical data

```{r}
load("mrr_bio.Rdata")
dim(clinical_data)
head(clinical_data[, 1:5])
colnames(clinical_data)
```

`clinical_data` is a table with 1231 rows and 24 columns:

- The rows correspond to the names of the patients, similar to `GeneX`.
- The columns are the names of the clinical data, such as `initial_weight`, `sites_of_involvement`, and `vital_status`.
- Each cell corresponds to the value of the clinical data for each patient, which can be either an integer or a character string.

### Outcome variable

The outcome variable is the vital status of the patient (dead or alive) :
```{r}
y <- clinical_data$vital_status
table(y)

barplot(table(y),
        main = "Distribution of Vital Status",
        xlab = "Vital Status",
        ylab = "Frequency",
        col = c("lightblue", "salmon"))
```


## Exploratory data analysis

### Genes

For the genes table, let's first plot for each gene the gaussian distribution of dead/alive for each gene. Then we can take the average of this graph.
And then plot all the averages corresponding to each gene in a common graph and check which ones vary strongly from the median (the median of the averages).

### Clinical Data

### Data cleaning

First model and confusion matrix :
```{r}
x <- clinical_data[, c("initial_weight",
                       "classification_of_tumor",
                       "morphology")]
y <- clinical_data$vital_status
model_data <- data.frame(x, vital_status = factor(y))
model_data <- na.omit(model_data) # removes NA for now
model_data$vital_status <- droplevels(model_data$vital_status)
```

First logistic model :
```{r}
model <- glm(vital_status ~ ., data = model_data, family = binomial)
```

Prediction : 
```{r}
pred <- predict(model, type = "response")
positive_class <- levels(model_data$vital_status)[2]
pred_class <- ifelse(pred > 0.5, positive_class, levels(model_data$vital_status)[1])
pred_class <- factor(pred_class, levels = levels(model_data$vital_status))
```

Confusion matrix :
```{r}
conf_mat <- table(Predicted = pred_class, Observed = model_data$vital_status)
conf_mat
```

Visualization of the confusion matrix
```{r}
conf_mat_matrix <- as.matrix(conf_mat)
image(t(conf_mat_matrix[nrow(conf_mat_matrix):1, ]),
  axes = FALSE,
  col = colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(64),
  xlab = "Observed",
  ylab = "Predicted",
  main = "Confusion Matrix")
axis(1, at = seq(0, 1, length.out = ncol(conf_mat_matrix)), labels = colnames(conf_mat_matrix))
axis(2, at = seq(0, 1, length.out = nrow(conf_mat_matrix)), labels = rev(rownames(conf_mat_matrix)))
text(rep(seq(0, 1, length.out = ncol(conf_mat_matrix)), each = nrow(conf_mat_matrix)),
     rep(seq(0, 1, length.out = nrow(conf_mat_matrix)), times = ncol(conf_mat_matrix)),
     labels = as.vector(conf_mat_matrix[rev(seq_len(nrow(conf_mat_matrix))), ]),
     col = "black", font = 2)
```

## Data analysis

### Genes

### Clinical data

#### Data cleaning

Here is each type corresponding to each data point as well as the frequencies of each occurrence.

After analysis we can remove redundant data like `age_at_diagnosis`, `days_to_birth` which are like the age in `age_at_index` and `bcr_patient_barcode` which is like the index of the line, and also `disease_type` and `primary_site` because they contain only one one observation which is the same fall all patients.

```{r}
redundant_cols <- c("age_at_diagnosis", "days_to_birth", "bcr_patient_barcode",
                    "disease_type", "primary_site")
clinical_data <- clinical_data[, setdiff(names(clinical_data), redundant_cols)]
cat("Removed columns:", paste(redundant_cols, collapse = ", "), "\n")
```

```{r, fig.width=6, fig.height=4, fig.show="asis"}
for (col in names(clinical_data)) {
  tbl <- table(as.character(clinical_data[[col]]), useNA = "ifany")
  
  plot.new() # start a fresh plot for each iteration
  
  barplot(tbl,
          main = paste("Distribution of", col),
          ylab = "Frequency",
          col = "steelblue",
          las = 2,
          cex.names = 0.7)
}
```