---
title: "mrr_project_bio"
author: ""
date: "2025-11-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Classification of Cancer outcome using Genetic and Clinical data

## Introduction

This project investigates the effect of genetic and clinical variables on the survival outcome of breast cancer patients. The data contains records of more than 1000 breast cancer patients from several research institutions. Clinical data contains patient-related and tumor-related information. Additionally mRNA gene expression data is available for each patient. The gene expression data has been processed to include only the top 5000 most variable genes on the transformed scale (log2(counts + 1)).

We consider that the main outcome variable of interest is **vital_status**, which is defined in clinical data.

## Load data

```{r}
load("mrr_bio.Rdata")
```

### Genes data

```{r}
dim(GeneX)
str(GeneX)
colnames(GeneX)[1:10]
rownames(GeneX)[1:10]
```

`GeneX` is a table with 1231 rows and 5000 columns:

- The rows correspond to the patients with each row name being the patient identifier such as `EW-A2FS-01A-11R-A17B`.
- The columns correspond to the genes with each column name being a gene such as `CLEC3A`, which are the 5000 most variable genes from one person to another.
- Each cell corresponds to the expression value (a float) of a certain gene (column) for a certain patient (row) -> continuous variables

### Clinical data

```{r}
load("mrr_bio.Rdata")
dim(clinical_data)
head(clinical_data[, 1:5])
colnames(clinical_data)
```

`clinical_data` is a table with 1231 rows and 24 columns:

- The rows correspond to the names of the patients, similar to `GeneX`.
- The columns are the names of the clinical data, such as `initial_weight`, `sites_of_involvement`, and `vital_status`.
- Each cell corresponds to the value of the clinical data for each patient, which can be either an integer or a character string.

### Outcome variable

The outcome variable is the vital status of the patient (dead or alive).
```{r}
y <- clinical_data$vital_status
table(y)

barplot(table(y),
        main = "Distribution of Vital Status",
        xlab = "Vital Status",
        ylab = "Frequency",
        col = c("lightblue", "salmon"))
```


## Exploratory data analysis

### Genes

For the genes table, let's first plot for each gene the gaussian distribution of proportion of dead minus proportion of alive and then take the average of this graph.

```{r}
# Calculate mean expression for each gene by vital status
for (col in colnames(GeneX)[1:5]) {  # First 5 genes for demonstration
  gene_data <- GeneX[, col]
  status <- clinical_data$vital_status
  
  # Split data by status
  alive_expr <- gene_data[status == "Alive"]
  dead_expr <- gene_data[status == "Dead"]
  
  # Calculate means
  mean_alive <- mean(alive_expr, na.rm = TRUE)
  mean_dead <- mean(dead_expr, na.rm = TRUE)
  
  # Create density plot
  plot(density(alive_expr, na.rm = TRUE), 
       col = "blue", 
       lwd = 2,
       main = paste("Gene:", col),
       xlab = "Expression Level",
       ylab = "Density")
  lines(density(dead_expr, na.rm = TRUE), 
        col = "red", 
        lwd = 2)
  
  # Add mean lines
  abline(v = mean_alive, col = "blue", lty = 2, lwd = 2)
  abline(v = mean_dead, col = "red", lty = 2, lwd = 2)
  
  # Add legend
  legend("topright", 
         legend = c(paste("Alive (mean =", round(mean_alive, 3), ")"),
                   paste("Dead (mean =", round(mean_dead, 3), ")")),
         col = c("blue", "red"), 
         lty = 1, 
         lwd = 2)
}
```

Thus we can display the most important genes by calculating the relative mean difference between the two groups (alive and dead) for each gene and then sorting them.

```{r}
gene_diff <- data.frame(
  Gene = colnames(GeneX),
  MeanDiff = numeric(ncol(GeneX))
)

for (i in seq_along(colnames(GeneX))) {
  col <- colnames(GeneX)[i]
  gene_data <- GeneX[, col]
  status <- factor(clinical_data$vital_status)
  
  plot_data <- data.frame(Expression = gene_data, Status = status)
  means <- aggregate(Expression ~ Status, data = plot_data, mean)
  
  # Calculate relative mean difference
  mean_alive <- means$Expression[means$Status == "Alive"]
  mean_dead <- means$Expression[means$Status == "Dead"]
  
  # Relative difference: (Dead - Alive) / Alive
  # Handle division by zero
  if (mean_alive != 0) {
    gene_diff$MeanDiff[i] <- abs((mean_dead - mean_alive) / mean_alive)
  } else {
    gene_diff$MeanDiff[i] <- NA
  }
}

# Remove genes with NA values
gene_diff <- gene_diff[!is.na(gene_diff$MeanDiff), ]

# Sort by relative mean difference
gene_diff <- gene_diff[order(gene_diff$MeanDiff, decreasing = TRUE), ]

# Display top 50 genes
top_50 <- head(gene_diff, 50)
print(top_50)

# Optional: visualize the distribution
barplot(top_50$MeanDiff[1:50],
        names.arg = top_50$Gene[1:50],
        las = 2,
        cex.names = 0.6,
        main = "Top 50 Genes by Relative Mean Difference",
        ylab = "Relative Mean Difference",
        col = "steelblue")
```

### Clinical Data

First modeling attempt with three variables from clinical data : `initial_weight`, `classification_of_tumor`, `morphology` to predict `vital_status`.
```{r}
x <- clinical_data[, c("initial_weight",
                       "classification_of_tumor",
                       "morphology")]
y <- clinical_data$vital_status
model_data <- data.frame(x, vital_status = factor(y))
model_data <- na.omit(model_data) # removes NA for now
model_data$vital_status <- droplevels(model_data$vital_status)
```

Model fitting:
```{r}
model <- glm(vital_status ~ ., data = model_data, family = binomial)
```

Prediction:
```{r}
pred <- predict(model, type = "response")
positive_class <- levels(model_data$vital_status)[2]
pred_class <- ifelse(pred > 0.5, positive_class, levels(model_data$vital_status)[1])
pred_class <- factor(pred_class, levels = levels(model_data$vital_status))
```

Confusion matrix:
```{r}
conf_mat <- table(Predicted = pred_class, Observed = model_data$vital_status)
conf_mat
```

Visualization of the confusion matrix.
```{r}
conf_mat_matrix <- as.matrix(conf_mat)
image(t(conf_mat_matrix[nrow(conf_mat_matrix):1, ]),
  axes = FALSE,
  col = colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(64),
  xlab = "Observed",
  ylab = "Predicted",
  main = "Confusion Matrix")
axis(1, at = seq(0, 1, length.out = ncol(conf_mat_matrix)), labels = colnames(conf_mat_matrix))
axis(2, at = seq(0, 1, length.out = nrow(conf_mat_matrix)), labels = rev(rownames(conf_mat_matrix)))
text(rep(seq(0, 1, length.out = ncol(conf_mat_matrix)), each = nrow(conf_mat_matrix)),
     rep(seq(0, 1, length.out = nrow(conf_mat_matrix)), times = ncol(conf_mat_matrix)),
     labels = as.vector(conf_mat_matrix[rev(seq_len(nrow(conf_mat_matrix))), ]),
     col = "black", font = 2)
```

## Data analysis

### Genes

Ce bloc pré-filtre les gènes par test univarié (top N) puis ajuste une unique régression logistique (glm) avec ces gènes simultanément. Il affiche le résumé, l'AUC et la matrice de confusion.

```{r simple-logistic-genes, echo=TRUE, warning=FALSE, message=FALSE}
# Paramètre : nombre de gènes à garder après pré-filtrage
N <- 50

# Outcome
y <- clinical_data$vital_status
if (!is.factor(y)) y <- factor(y)
stopifnot(nrow(GeneX) == length(y))

# Pré-filtrage univarié (t-test p-values)
pv <- apply(GeneX, 2, function(g) {
  g1 <- g[y == levels(y)[1]]
  g2 <- g[y == levels(y)[2]]
  if (length(na.omit(g1)) < 2 || length(na.omit(g2)) < 2) return(1)
  res <- try(t.test(g1, g2)$p.value, silent = TRUE)
  if (inherits(res, "try-error")) return(1)
  res
})
pv[is.na(pv)] <- 1

# Top N gènes
top_genes <- names(sort(pv, decreasing = FALSE))[1:min(N, length(pv))]

# Construire jeu de données et ajuster glm (logistic)
model_df <- data.frame(vital_status = y, as.data.frame(GeneX[, top_genes, drop = FALSE]))
model_df <- na.omit(model_df)
model_df$vital_status <- droplevels(model_df$vital_status)

glm_fit <- glm(vital_status ~ ., data = model_df, family = binomial)

# Résumé
cat('\n--- Résumé du modèle (GLM) ---\n')
print(summary(glm_fit))

pred_prob <- predict(glm_fit, type = 'response')

# Sauvegarder la liste des top genes utilisés
write.csv(data.frame(top_genes = top_genes), file = 'top_genes_used_in_glm.csv', row.names = FALSE)
 
## Graphe des classes (TP/TN en vert, FP/FN en orange)
# Convert probabilities to predicted class
# with 0 being below the median and 1 above
pred_resp <- pred_prob
# construire y binaire (0/1) compatible
y <- ifelse(model_df$vital_status == levels(model_df$vital_status)[2], 1, 0)
pred_class <- ifelse(pred_resp >= 0.5, 1, 0)
point_colors <- ifelse(
  pred_class == y,
  "green", # True positives / True negatives
  "orange" # False positives / False negatives
)

plot(
  y + runif(length(y), -0.02, 0.02),
  pred_resp,
  xlab = "Actual y",
  ylab = "Predicted probability",
  main = "Predicted vs Actual (Green = TP/TN, Orange = FP/FN)",
  col = point_colors,
  pch = 19,
  cex = 0.5
)

x_log <- seq(0, 1, length.out = 200)
y_log <- 1 / (1 + exp(-10 * (x_log - 0.5)))
lines(x_log, y_log, col = "red", lwd = 2)
abline(h = 0.5, col = "blue", lty = 2)
```


```{r}
## Visualisation de la matrice de confusion (heatmap)
# Recalculer conf_mat pour avoir Predicted vs Observed
conf_mat_final <- table(Predicted = pred_class, Observed = model_df$vital_status)

# Indicateurs simples
accuracy <- sum(diag(conf_mat_final)) / sum(conf_mat_final)
cat(sprintf("\nAccuracy: %.3f\n", accuracy))

# Visualisation de la matrice de confusion
conf_mat_matrix <- as.matrix(conf_mat_final)
image(t(conf_mat_matrix[nrow(conf_mat_matrix):1, ]),
  axes = FALSE,
  col = colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(64),
  xlab = "Observed",
  ylab = "Predicted",
  main = "Confusion Matrix")
axis(1, at = seq(0, 1, length.out = ncol(conf_mat_matrix)), labels = colnames(conf_mat_matrix))
axis(2, at = seq(0, 1, length.out = nrow(conf_mat_matrix)), labels = c("Dead", "Alive"))
text(rep(seq(0, 1, length.out = ncol(conf_mat_matrix)), each = nrow(conf_mat_matrix)),
     rep(seq(0, 1, length.out = nrow(conf_mat_matrix)), times = ncol(conf_mat_matrix)),
     labels = as.vector(conf_mat_matrix[rev(seq_len(nrow(conf_mat_matrix))), ]),
     col = "black", font = 2)
```

### Clinical data

#### Data cleaning

Here is each type corresponding to each data point as well as the frequencies of each occurrence.

After analysis we can remove redundant data like `age_at_diagnosis`, `days_to_birth` which are like the age in `age_at_index` and `bcr_patient_barcode` which is like the index of the line, and also `disease_type` and `primary_site` because they contain only one one observation which is the same fall all patients.

```{r}
redundant_cols <- c("age_at_diagnosis", "days_to_birth", "bcr_patient_barcode",
                    "disease_type", "primary_site")
clinical_data <- clinical_data[, setdiff(names(clinical_data), redundant_cols)]
cat("Removed columns:", paste(redundant_cols, collapse = ", "), "\n")
```

```{r, fig.width=6, fig.height=4, fig.show="asis"}
for (col in names(clinical_data)) {
  tbl <- table(as.character(clinical_data[[col]]), useNA = "ifany")
  
  plot.new() # start a fresh plot for each iteration
  
  barplot(tbl,
          main = paste("Distribution of", col),
          ylab = "Frequency",
          col = "steelblue",
          las = 2,
          cex.names = 0.7)
}
```

#### Encoding categorical variables

Label encoding of the `clinical_data` categorical variables, with NA also as a category. Label encoding (bijection) for non-integer clinical columns :
```{r}
label_encoded_clinical <- clinical_data
label_mappings <- list()

for (col in names(label_encoded_clinical)) {
  column_values <- label_encoded_clinical[[col]]
  if (!is.numeric(column_values)) {
    if (!is.atomic(column_values)) {
      column_values <- vapply(column_values, function(entry) {
        if (is.null(entry) || (length(entry) == 1 && is.na(entry))) {
          return(NA_character_)
        }
        paste0(as.character(entry), collapse = "|")
      }, character(1))
    }
    factor_values <- factor(column_values, exclude = NULL)
    label_encoded_clinical[[col]] <- as.integer(factor_values)
    label_mappings[[col]] <- levels(factor_values)
  }
}

label_encoded_clinical[] <- lapply(label_encoded_clinical, function(col) {
  if (is.numeric(col)) {
    col[is.na(col)] <- -1 # NA placeholder for numeric columns
  } else {
    f <- factor(col, exclude = NULL)
    col <- as.integer(f)
  }
  col
})

cat("Label-encoded clinical data dimensions:",
    paste(dim(label_encoded_clinical), collapse = " x "), "\n")

if (length(label_mappings) > 0) {
  cat("Bijections:\n")
  for (col in names(label_mappings)) {
    cat(col, "=>",
        paste(seq_along(label_mappings[[col]]),
              label_mappings[[col]],
              sep = ":", collapse = ", "),
        "\n")
  }
}
```

#### Logistic RIDGE Regression

```{r}
library(glmnet)

predictors <- label_encoded_clinical
predictors$vital_status <- NULL

# Check for NA values in predictors and prints them and Check for NA values in clinical_data$vital_status and prints them
if (any(is.na(predictors))) {
  na_rows <- which(rowSums(is.na(predictors)) > 0)
  cat("Rows with NA in predictors:", na_rows, "\n")
}
if (any(is.na(clinical_data$vital_status))) {
  na_rows_vital <- which(is.na(clinical_data$vital_status))
  cat("Rows with NA in vital_status:", na_rows_vital, "\n")
}

# Remove rows with NA in vital_status
model_data <- na.omit(data.frame(predictors, vital_status = clinical_data$vital_status))

# Prepare predictor matrix X (numeric matrix) and response y (factor) for glmnet
X <- as.matrix(model_data[, setdiff(colnames(model_data), "vital_status"), drop = FALSE])
y <- factor(model_data$vital_status)

# Fit a logistic ridge regression model using cross-validation
ridge_fit <- cv.glmnet(X, y, family = "binomial", alpha = 0)
best_lambda <- ridge_fit$lambda.min

# Performs predictions using logistic ridge regression and evaluates the model's performance.
ridge_pred_prob <- predict(ridge_fit, newx = X, s = "lambda.min", type = "response")
threshold <- 0.5
predicted_class <- ifelse(ridge_pred_prob > threshold, levels(y)[2], levels(y)[1])
ridge_conf_mat <- table(Predicted = predicted_class, Observed = y)
ridge_conf_mat

# Compute and print ridge confusion matrix accuracy
ridge_accuracy <- sum(diag(ridge_conf_mat)) / sum(ridge_conf_mat)
cat("Ridge accuracy:", round(ridge_accuracy, 3), "\n")

# Visualization of the ridge confusion matrix
ridge_conf_mat_matrix <- as.matrix(ridge_conf_mat)
image(t(ridge_conf_mat_matrix[nrow(ridge_conf_mat_matrix):1, ]),
   axes = FALSE,
   col = colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(64),
   xlab = "Observed",
   ylab = "Predicted",
   main = "Ridge Confusion Matrix")
axis(1,
  at = seq(0, 1, length.out = ncol(ridge_conf_mat_matrix)),
  labels = colnames(ridge_conf_mat_matrix))
axis(2,
  at = seq(0, 1, length.out = nrow(ridge_conf_mat_matrix)),
  labels = rev(rownames(ridge_conf_mat_matrix)))
text(rep(seq(0, 1, length.out = ncol(ridge_conf_mat_matrix)), each = nrow(ridge_conf_mat_matrix)),
  rep(seq(0, 1, length.out = nrow(ridge_conf_mat_matrix)), times = ncol(ridge_conf_mat_matrix)),
  labels = as.vector(ridge_conf_mat_matrix[rev(seq_len(nrow(ridge_conf_mat_matrix))), ]),
  col = "black", font = 2)
```


### Clinical data with Genes


## Extension